\chapter*{Acknowledgements}
\addcontentsline{toc}{chapter}{Acknowledgements}
\markboth{Acknowledgements}{Acknowledgements}

On October 1st, 2018, I started my PhD research. The work done in the following years eventually led to this thesis. I attribute this to a trifecta of excellent supervision, collaboration, and support. For this, I would like to thank some people:

First and foremost, thank you, Arjen. Thank you for the opportunity to be your student. Throughout the years, I have learned a lot from you. I have always enjoyed working together with you, and I still do. I will always be appreciative of what you have offered me. 

I want to thank the manuscript committee for reading my thesis carefully and providing me with great feedback. Your input really helped me. 

I am grateful for the collaborations that helped my research. Jimmy, I really enjoyed working together on a couple of occasions. I liked seeing how you approach research projects. Working with you showed me how I can approach work more structured, helping me become more efficient in my work. Faegheh, our collaborations were a pleasure as well. I have learned a lot from you working together. You have a great attention to detail, without losing sight of the bigger picture.

When I started my research, Arjen and I formed a two-person Information Retrieval research group. Over the years, this changed to a large group of Information Retrieval researchers. Emma, you were the first person to join the group, and we have been office mates throughout our appointments as PhD students. We were in the same situation, making it possible to always share our work experiences with someone going through the same. The Information Retrieval research group grew rapidly, making the job even more delightful. So thank you, Djoerd, Martha, Faegheh, Harrie, Ivan, Koen, Neghin, Hideaki, Norman, Tim, Gijs, Daria, Mohanna, Heydar, Thomas, and Mick.

Our research group was, of course, part of a larger group of researchers in the department of data science at the institute of computing science. All colleagues at DAS were very kind and knowledgeable. I really enjoyed all our lunch sessions together, as well as all board game nights and other informal events. I am very lucky that I was able to work in such a great environment. 

Then, I would say thank you to my current colleagues at Spinque. I really enjoy working at a company with so many great people. At Spinque, I can work on projects where I can use the things I learned during my PhD. Thank you also for your support when finishing my thesis. 

Mart, Koen, and Pieter, thank you for being great friends not only during my PhD research, but already throughout the bachelor's and master's programs. Mart, we have known each other even longer. Thank you for being such a good friend for such a long time. To all my other friends, thank you for being my friend. I am very lucky to have you all in my life. 
\subsubsection*{}
Mijn ouders wil ik bedanken voor een warm en fijn thuis. Op een goed fundament kun je bouwen, en die hebben jullie zeker gelegd. Stijn en Anne, ik ben heel blij met jullie als broer en zus. Jullie inspireren mij dagelijks. Ook mijn schoonfamilie wil ik bedanken, jullie staan altijd voor mij klaar.

Ten slotte, Maudy, bedankt voor jouw onuitputtelijke steun en liefde. Joep, de afgelopen maanden met jou waren geweldig. Ik kijk er naar uit om de rest van het leven samen met jullie door te brengen. 
