\chapter{Conclusion}
\label{conclusion}
\epigraph{We can only see a short distance ahead, but we can see plenty there that needs to be done.}{Alan Turing - 1950}

To conclude I would like to summarize the work described in the chapters following the research questions of this thesis. We will look one time more to the work in the individual chapters, and then try to tie it together following the main research question. Then finally, I want to reflect on the work described in this thesis and present research opportunities that follow the work described. 

\section{Contributions}
Recall the main research question, and its subquestions:
\begin{itemize}
	\item \textbf{Research Question: How can information retrieval benefit from graph databases and graph query languages?}
	\item \emph{Research Question 1: What are the benefits of using relational databases for information retrieval?} 
	\item \emph{Research Question 2: Can we extend the benefits from using relational databases for information retrieval to using graph databases, while being able to express graph related problems easier?} 
	\item \emph{Research Question 3: When does information retrieval research benefits from graph data?} 
\end{itemize}

\cref{ir-using-relational-databases} introduces the history of using relational databases for information retrieval, one of the latter attempts, using columnar databases for retrieval is highlighted. This approach is re-implemented as a prototype system; ``OldDog''. OldDog is used for a reproduction experiments, where many implementations of BM25 are compared against each other. Because OldDog was built using a relational database, we were able to express the logical model of the ranking functions separately from their physical models. This way we could exactly find what the difference were between the models, while keeping the data exactly the same. The work in the chapter demonstrates the usefulness of using relational databases for reproduction experiments in information retrieval. Looking at research question 1: \emph{What are the benefits of using relational databases for information retrieval?}, we have demonstrated that relational databases provide a framework for easily comparing different ranking methods, as shown in the reproduction study. The systems is reasonably efficient, making it easy to set up for new experiments. 

\cref{from-tables-to-graphs} takes the data model for information retrieval using relational databases, as presented in~\cref{ir-using-relational-databases}, and extends this model to a graph data model. This model is implemented in the prototype system GeeseDB, a graph query language method for information retrieval. GeeseDB is built on top of DuckDB, and makes use of its column oriented tables and the fact that it can be ran in-process. By having a powerful backend, we are able to express graph queries for information retrieval and run them on GeeseDB. GeeseDB supports all functionalities that were already supported by OldDog, plus it makes the expression of more complicated problems, through the graph framework easier. Considering research question 2: \emph{Can we extend the benefits from using relational databases for information retrieval to using graph databases, while being able to express graph related problems easier?}, we find that, with GeeseDB, it is possible to take all the benefits from using relational databases for information retrieval, while making it possible to express more complex problems through a graph query language.

\cref{a-graph-of-entities} concerns the improvements on the Radboud Entity Linking (REL) system. When trying to deploy this system in order to annotate a large document collection, some unforeseen issues were encountered. The time it took to tag the whole collection was much larger than expected, making it unfeasible to annotate the whole corpus in a reasonable time. By improving the efficiency of several components of the software, and including a batch extension, it was possible to speed up parts of the software by a factor of ten. After deployment of these improvements, it was possible to use the REL software for large corpora. 

\cref{ch:mmead} continues with the system created in \cref{a-graph-of-entities} and deploys it to annotate the large MS MARCO document and passage collections. For these annotations a specification on how to share them, independent of linking system, is proposed. Following this specification we make the annotations created by the REL system publicly available. We also publish data created by the BLINK entity linking system for the MS MARCO v1 passage collection. Using the annotations by REL, we show on the MS MARCO v1 collection that query expansion with entity annotations significantly improves recall for complicated queries. Although this is just a simple experiment, we show that these annotations contain information useful for finding relevant documents, and they could be beneficial for more sophisticated methods. Also, through a demonstration that combines entity annotations with geographical information, we show that entity annotations benefit interactive search applications. Considering research question 3: \emph{When does information retrieval research benefits from graph data?}. We demonstrate that information retrieval can benefit from graph data, in this case an entity graph. By employing it for retrieval techniques it leads to an increase in retrieval effectiveness. Also, the graph model is ideal for interactive search systems. 

Finally, to address the main research question; \textbf{How can information retrieval benefit from graph databases and graph query languages?}. This thesis demonstrates the usefulness of graph databases for information retrieval by creating a prototype system, GeeseDB, that is directly useful for information retrieval. With GeeseDB, it is possible to set up retrieval experiments quickly that make use of the graph model. It inherits all benefits of using relational databases for information retrieval. We show that using a graph of entities we can increase retrieval effectiveness. Moreover, this graph of entities can improve interactive search applications as well.    

\section{Future Work}
Following the work presented in this thesis, there are interesting research opportunities that can follow from this work. In particular I would like to highlight three of them:
\begin{enumerate}
	\item Firstly, an obvious follow up on the work presented in this thesis is a study that combines the work of~\cref{from-tables-to-graphs,ch:mmead}. In~\cref{from-tables-to-graphs} we introduced GeeseDB, a graph based system for information retrieval. In~\cref{ch:mmead} we use graph data produced by the REL system. It would be interesting to see if we can reproduce the results of~\cref{ch:mmead} using GeeseDB. For the demonstration in~\cref{ch:mmead} we have used a RDF engine. Although this was a suited system for our demonstration, using our own software would have demonstrated additional benefits of using GeeseDB. 
	\item Then, the examples shown in~\cref{from-tables-to-graphs} are relatively simple. Additional studies should be carried out that take into account more complex graph structures. For example, can we employ a graph structure and utilize paths in the graph? These paths can for example be used to propagate probabilities that are used to calculate relevancy scores, a strategy that is probably hard to express without using graphs. 
	\item Lastly, it would be interesting to see if the concepts of graph databases for information retrieval and deep learning for information retrieval can be married together. In the last couple of years, deep learning for information retrieval has become more prominent, especially with the rise of large language models. I expect that the representations found by these models can expressed in graph databases, and that both techniques can benefit from each other. Additional studies need to be carried out to find out how this can be done well. 
\end{enumerate} 